The effect of linking assumptions and number of response options on inferred scalar implicature rate

### 2018-05-02 writeup.tex

Comparing the strong and weak rows on Figure
\ref{fig:implicatureRatePlot}, we see that a weak linking assumption
tends to estimate higher implicature rates, especially in tasks with
more response options. With a strong linking assumption, the binary and
possibly ternary judgment tasks derive higher implicature rates than
quaternary and quinary tasks. With a weak linking assumption, the
pattern is reversed. Quaternary and quinary tasks estimate higher rates
than binary and ternary tasks.

@@

We test this hypothesis formally using a binomial mixed effects model
with the fixed effect of response type and the random intercept for
participants as well as random intercept and slope for

@@

\section{Discussion}\label{discussion}

Alternative Linking Hypothesis: * RSA: Response behavior across
conditions (utterance-card combinations) and dependent measures can be
predicted by a linking hypothesis that assumes that participants are
behaving like soft-optimal RSA speakers and provide a particular
response (eg TRUE) to an utterance u if the RSA speaker probability of u
(given the card) is within a particular probability interval (eg, within
the interval {[}theta, 1{]}).

\begin{itemize}
\tightlist
\item
  Differences between traditional approaches and RSA: 1. The traditional
  linking hypotheses are based on a binary implicature/literal theory of
  pragmatic reasoning but RSA gives a continuous measure of pragmatic
  reasoning and allows for better predicting response behavior with
  multiple options.
\end{itemize}

A simple visual inspection of our results suggests that any variant of
the Traditional Linking Hypothesis will yield unsatisfying empirical
coverage with regards to participants' behavior in truth value judgment
tasks. Recall that according to this hypothesis, scalar implicature is
conceptualized as a binary, categorical affair: an implicature is either
\enquote{calculated,} or it isn't. This assumption has implications for
how we approach analysis of variation in behavior on a truth value
judgment task; for example, why did the majority of respondents in the
binary condition of our experiment answer \enquote{Right} to an
utterance of cat or dog when the card had a cat and a dog on it?

On the Traditional story, we are forced to say that a) not all
participants calculated the implicature; or that b) some participants
who calculated the implicature did not choose the anticipated response
(i.e. \enquote{Wrong}) due to some other cognitive reflex which
\enquote{overrode} the implicature; or some mixture of (b) and (c). We
might similarly posit that one or both of these factors underlie the
variation in the ternary, quatenary, and quinary conditions (e.g.~why
were participants roughly split between \enquote{Right} and
\enquote{Kind of right} when the utterance was cat or dog and the card
had a cat and a dog?). There is an established cottage industry of
pragmatics research which approaches the problem of behavioral variation
in exactly these ways, exploring the factors which \enquote{license}
implicature calculation or \enquote{suppress} its anticipated behavioral
expression (c.f. \_\_), but the best we can hope for on this approach is
an analysis which traces the general qualitative patterns in the data
(and we should stress that even in this respect, the industry has yet to
fully deliver the goods!).

The Alternative Linking Hypothesis contrasts with the Traditional
Linking Hypothesis in that the former is rooted in a quantitative
formalization of pragmatic competence which provides us a continuous
measure of pragmatic reasoning. {[}recap RSA model here? Maybe show how
it gets the right basic qualitative pattern, i.e. \(P_L\)(a and
b\textbar{} \enquote{a or b}) \(<\) \(P_L\)(a \textbar{} \enquote{or});
\(P_L\)(a and b\textbar{} \enquote{a or b}) \(<\) \(P_L\)(b \textbar{}
\enquote{or}) {]}

Following Degen \& Goodman (2014), we proceed on the assumption that
behavior on sentence verification tasks, such as truth value judgment
tasks, is best modeled as a function of an individual's mental
representation of a cooperative interlocutor (\(S_1\) in the language of
RSA). For a given utterance u and an intended communicated meaning w,
\(S_1\)(u \textbar{} w) outputs a conditional probability of u given w.
For example, in the binary condition of our experiment where a
participant evaluated cat or dog when there were both animals on the
card, the participant has access to the mental representation of \(S_1\)
and hence to the \(S_1\) conditional probability of hearing the
utterance cat or dog given a dog and cat card \(S_1\)(cat or dog
\textbar{} cat and dog).

One desideratum of a linking hypothesis given this view of pragmatic
competence is to transform this probability value into a categorical
output (e.g. \enquote{Right}/''Wrong'' in the case of the binary
condition). This can be achieved by positing that participants
\enquote{filter out} low-probability utterances given an intended
communicated meaning. We model a responder, R, who in the binary
condition responds \enquote{Right} to an utterance u in world w just in
case \(S_1\)(u \textbar{} w) exceeds some probability threshold
\(\theta\):

R(u, w, \(\theta\)) = \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\)
\(\theta\) = \enquote{Wrong} otherwise

In the experiment conditions where there are more than two choices, we
model the range of possible behavioral responses for R with the
introduction of intermediate probability thresholds. For example, in the
ternary condition, R(u, w, \(\theta_1\) , \(\theta_2\)) is
\enquote{Right} iff \(S_1\)(u \textbar{} w) \textgreater{} \(\theta_1\)
and \enquote{Neither} iff \(\theta_1\) \textgreater{} \(S_1\)(u
\textbar{} w) \textgreater{} \(\theta_2\). To fully generalize the model
to our five experimental conditions, we say that R takes as its input an
utterance u and a world state w and a number of threshold variables
dependent on a variable c, corresponding to the experimental condition
in which the participant finds herself (e.g.~the range of possible
responses available to R).

@@

R(u, w, \(\theta_1\) , \(\theta_2\))\\
= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\) =
\enquote{Neither} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w) \(>\)
\(\theta_2\)\\
= \enquote{Wrong} otherwise

@@

R(u, w, \(\theta_1\) , \(\theta_2\), \(\theta_3\))\\
= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\) =
\enquote{Kinda Right} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_2\)\\
= \enquote{Kinda Wrong} iff \(\theta_2\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_3\) = \enquote{Wrong} otherwise

@@

R(u, w, \(\theta_1\) , \(\theta_2\), \(\theta_3\). \(\theta_4\))\\
= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\)
=\enquote{Kinda Right} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_2\) = \enquote{Neither} iff \(\theta_2\) \(>\) \(S_1\)(u
\textbar{} w) \(>\) \(\theta_3\)\\
= \enquote{Kinda Wrong} iff \(\theta_3\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_4\) = \enquote{Wrong} otherwise

\newpage

>>>>

Comparing the strong and weak rows on Figure
\ref{fig:implicatureRatePlot}, we see that a weak linking assumption
tends to estimate higher implicature rates, especially in tasks with
more response options. With a strong linking assumption, the binary and
possibly ternary judgment tasks derive higher implicature rates than
quaternary and quinary tasks. With a weak linking assumption, the
pattern is reversed. Quaternary and quinary tasks estimate higher rates
than binary and ternary tasks. The patterns show that estimates of
\enquote{implicature rate} depend on linking assumptions and the number
of responses available to participants in the study.

Comparing the exhaustive and scalar columns of Figure
\ref{fig:implicatureRatePlot}, we see that with a strong linking
assumption, there are slightly higher rates for scalar implicatures in
the binary and ternary tasks. With a weak linking assumption, there may
be slightly higher rates for scalar implicatures in the binary and
ternary while the rates may be lower in the quaternary and quinary
tasks. In what follows, we test the effect of linking assumption and
response options on exhaustive and scalar implicature rates more
formally.

@@

For our formal analysis, we use a binomial mixed effects model with the
fixed effect of response type (binary, ternary, quaternary, quinary),
linking assumption (strong vs.~weak), and trial type (exhaustive
vs.~scalar), as wellas the random intercept for participants as well as
random intercept and slope for items (the card they saw).

@@

\section{\# Discussion}\label{discussion}

On the traditional view of the link between implicature and behavior in
sentence verification tasks, scalar implicature is conceptualized as a
binary, categorical affair - that is, an implicature is either
\enquote{calculated} or it isn't, and the behavioral reflexes of this
categorical interpretation process should be straightforwardly observed
in experimental paradigms. This assumption has concerning implications
for how we must approach analysis of variation in behavior on a truth
value judgment task; for example, why did the majority of respondents in
the binary condition of our experiment answer \enquote{Right} to an
utterance of cat or dog when the card had both a cat and a dog on it?

To explain the data on the traditional view, are forced to say that a)
not all participants calculated the implicature; or that b) some
participants who calculated the implicature did not choose the
anticipated response (i.e. \enquote{Wrong}) due to some other cognitive
reflex which \enquote{overrode} the implicature; or some mixture of (b)
and (c). We might similarly posit that one or both of these factors
underlie the variation in the ternary, quatenary, and quinary conditions
(e.g.~why were participants roughly split between \enquote{Right} and
\enquote{Kind of right} when the utterance was cat or dog and the card
had a cat and a dog?). However, the best we can hope for on this
approach is an analysis which traces the general qualitative patterns in
the data.

We contrast the above view of implicature and its behavioral reflexes
with an alternative linking hypothesis which assumes that participants'
behavior can be represented using the model of a soft-optimal pragmatic
speaker in the RSA framework. This alternative linking hypothesis
contrasts with the traditional view in it is rooted in a quantitative
formalization of pragmatic competence which provides us a continuous
measure of pragmatic reasoning. Recall that in RSA, pragmatically
competent listeners are modeled as a continuous probabilistic
distribution of possible meanings given an utterance which that listener
hears. The probability with which this listener \(L_1\) ascribes a
meaning s to an utterance u depends upon a prior probability
distribution of potential states of the world \(P_w\), and upon
reasoning about the communicative behavior of a speaker \(S_1\). \(S_1\)
in turn is modeled as a continuous probabilistic distribution of
possible utterances given an intended state of affairs the speaker
intends to communicate. This distribution is sensitive to a rationality
parameter \(\alpha\), the production cost \(C\) of potential utterances,
and a representation of a literal listener \(L_0\) whose interpretation
of an utterance is in turn a function of that utterance's truth
conditional content \([[u]](s)\) and her prior beliefs about the state
of the world \(P_w(s)\).

\begin{verbatim}
$P_{L_1}(s | u) \propto P_{S_1}(u | s) * P_w(s)$

$P_{S_1}(u | s) \propto exp(\alpha(log(L_0(s | u)) - C(u))) $

$P_{L_0}(s | u) \propto [[u]](s) * P_w(s)$
\end{verbatim}

In this framework, individuals never categorically draw (or fail to
draw) pragmatic inferences about the utterances they hear. For example,
exclusivity readings of disjunction or are represented in RSA as
relatively low conditional probability of a conjunctive meaning on the
\(P_L\) distribution, given an utterance of or. Thus, it is not even
possible to talk about \enquote{rate} of implicature calculation in the
RSA framework. The upshot, as we show below, is that this view of
pragmatic competence does allow us to talk explicitly and quantitatively
about rates of observed behavior in sentence verification tasks.

First, following Degen \& Goodman (2014), we proceed on the assumption
that behavior on sentence verification tasks, such as truth value
judgment tasks, is best modeled as a function of an individual's mental
representation of a cooperative interlocutor (\(S_1\) in the language of
RSA) rather than of a pragmatic listener who interprets utterances
(\(P_{L_1}\)). In their paper, Degen \& Goodman argue that sentence
verification tasks are relatively more sensitive to contextual
manipulations (such as manipulation of the Question Under Discussion)
than are sentence interpretation tasks, and that this follows if
sentence interpretation tasks - but not sentence verification tasks -
require an additional layer of counterfactual reasoning about the
intentions of a cooperative speaker.

A main desideratum of a behavioral linking hypothesis given the RSA view
of pragmatic competence is to transform continuous probability
distributions into categorical outputs (e.g.~responses of
\enquote{Right}/''Wrong'' in the case of the binary condition of our
experiment). For a given utterance u and an intended communicated
meaning w, \(S_1\)(u \textbar{} w) outputs a conditional probability of
u given w. For example, in the binary condition of our experiment where
a participant evaluated cat or dog when there were both animals on the
card, the participant has access to the mental representation of \(S_1\)
and hence to the \(S_1\) conditional probability of hearing the
utterance cat or dog given a dog and cat card: \(S_1\)(cat or dog
\textbar{} cat and dog). According to the linking hypothesis advanced
here, the participant provides a particular response to u if the RSA
speaker probability of u lies within a particular probability interval,
given an observed state of the world (i.e.~the configuration of animals
on the card in our experiment). We model a responder, R, who in the
binary condition responds \enquote{Right} to an utterance u in world w
just in case \(S_1\)(u \textbar{} w) exceeds some probability threshold
\(\theta\):

R(u, w, \(\theta\))

= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta\)

= \enquote{Wrong} otherwise

In the experiment conditions where there are more than two choices, we
model the range of possible behavioral responses for R with the
introduction of intermediate probability thresholds. For example, in the
ternary condition, R(u, w, \(\theta_1\) , \(\theta_2\)) is
\enquote{Right} iff \(S_1\)(u \textbar{} w) \textgreater{} \(\theta_1\)
and \enquote{Neither} iff \(\theta_1\) \textgreater{} \(S_1\)(u
\textbar{} w) \textgreater{} \(\theta_2\). To fully generalize the model
to our five experimental conditions, we say that R takes as its input an
utterance u, a world state w, and a number of threshold variables
dependent on a variable c, corresponding to the experimental condition
in which the participant finds herself (e.g.~the range of possible
responses available to R).

@@

R(u, w, \(\theta_1\) , \(\theta_2\))

= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\)

= \enquote{Neither} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w) \(>\)
\(\theta_2\)

= \enquote{Wrong} otherwise

@@

R(u, w, \(\theta_1\) , \(\theta_2\), \(\theta_3\))

= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\)

= \enquote{Kinda Right} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_2\)

= \enquote{Kinda Wrong} iff \(\theta_2\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_3\)

= \enquote{Wrong} otherwise

@@

R(u, w, \(\theta_1\) , \(\theta_2\), \(\theta_3\). \(\theta_4\))

= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\)

=\enquote{Kinda Right} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_2\)

= \enquote{Neither} iff \(\theta_2\) \(>\) \(S_1\)(u \textbar{} w) \(>\)
\(\theta_3\)

= \enquote{Kinda Wrong} iff \(\theta_3\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_4\)

= \enquote{Wrong} otherwise

Bayesian statistical methods provide us a means for estimating the
values of these probability thresholds in our RSA model. The basis for
the model is a set of possible states of the world, given a universe of
three animals - X, Y, and Z - that each may be on some card. We next
define a set of possible sentences a speaker might utter, assuming the
speaker intends to communicate which animals are on the card. We assume
a uniform prior probability of different states of the world and a
uniform cost function on utterances. We define a literal listener
\(L_0\), a pragmatic speaker \(S_1\), and a responder \(R\) according to
our definitions above. Lastly, and assuming a uniform prior distribution
over possible values of probability thresholds, we use Bayesian
inference to recover a posterior distribution of these thresholds in
each experimental condition, given the actual observed rate of response
in each condition of the experiment. The results of this parameter
estimation analysis are shown in the figures below, where the X axis of
each figure corresponds to a threshold value between 0 and 1 and the Y
axis corresponds to the posterior probability density of possible values
of the threshold.

\begin{verbatim}
## Warning in is.na(e1) | is.na(e2): longer object length is not a multiple of
## shorter object length
\end{verbatim}

\begin{verbatim}
## Warning in `==.default`(Parameter, c("quatenary_theta1",
## "quatenary_theta2", : longer object length is not a multiple of shorter
## object length
\end{verbatim}

\includegraphics{writeup_files/figure-latex/unnamed-chunk-1-1.pdf}
\includegraphics{writeup_files/figure-latex/unnamed-chunk-1-2.pdf}
\includegraphics{writeup_files/figure-latex/unnamed-chunk-1-3.pdf}
\includegraphics{writeup_files/figure-latex/unnamed-chunk-1-4.pdf}

The above analysis is a proof of concept for the following idea: by
relaxing the assumptions of the traditional view of scalar implicature
(namely, that scalar implicatures either are or are not calculated, and
that behavior on sentence verification tasks directly reflects this
binary interpretation process), we can propose quantitative models of
the variation in behavior we observe in experimental settings.
